---
title: |
  | \huge A Brief Introduction to
  | \huge Programming with R - Exercises
subtitle: "MEcon & MiQE/F Introductory Week"
author: "Ana Armendariz & Federica Mascolo"
date: "September 4, 2026"
format:
  beamer:
    theme: default
    colortheme: default
    fonttheme: default
    navigation: horizontal
    section-titles: true
    toc: false
    slide-level: 3
    fig-width: 5
    fig-height: 3
    fig-align: center
    df-print: kable
    fontsize: 12pt
    aspectratio: 169
    keep-tex: false
    latex-engine: xelatex
    include-in-header:
      text: |
        \usepackage{fontspec}K
        \usepackage{tikz}
        \usetikzlibrary{shapes,arrows,positioning}

        % Colors
        \definecolor{mDarkBlue}{HTML}{23373b}
        \definecolor{mLightBlue}{HTML}{0066CC}
        \definecolor{mAccent}{HTML}{EB811B}
        \definecolor{hsgDarkGreen}{HTML}{006B54}
        \definecolor{mGray}{HTML}{4A4A4A}
        \definecolor{mLightGray}{HTML}{F5F5F5}
        \definecolor{mBackground}{HTML}{FAFAFA}
        \definecolor{mGreen}{HTML}{14B03D}
        \definecolor{mPurple}{HTML}{6B3AA0}

        % Fonts (fallbacks)
        \IfFontExistsTF{Fira Sans}{\setsansfont{Fira Sans}}{%
          \IfFontExistsTF{Helvetica Neue}{\setsansfont{Helvetica Neue}}{\setsansfont{Arial}}}

        % Body text size
        \setbeamerfont{normal text}{size=\normalsize}
        % All slide titles (##) as \large
        \setbeamerfont{frametitle}{size=\large,series=\bfseries}
        \setbeamerfont{section title}{size=\huge,series=\bfseries}
        \setbeamerfont{subsection title}{size=\large,series=\bfseries}

        % Beamer colors
        \setbeamercolor{normal text}{fg=mDarkBlue}
        \setbeamercolor{alerted text}{fg=mAccent}
        \setbeamercolor{example text}{fg=mLightBlue}
        \setbeamercolor{structure}{fg=mDarkBlue}
        \setbeamercolor{title}{fg=white,bg=mDarkBlue}
        \setbeamercolor{frametitle}{fg=mDarkBlue,bg=mBackground}
        \setbeamercolor{section in toc}{fg=mDarkBlue}
        \setbeamercolor{subsection in toc}{fg=mGray}
        \setbeamertemplate{navigation symbols}{}

        % Frame title with rule
        \setbeamertemplate{frametitle}{
          \vspace{0.4em}\insertframetitle\vspace{0em}
          \textcolor{mAccent}{\rule{\textwidth}{2pt}}
        }

        % Section page
        \AtBeginSection[]{
          \begin{frame}[plain]
            \vfill\centering
            \begin{beamercolorbox}[sep=12pt,center,shadow=false,rounded=true]{title}
              {\Huge\bfseries\insertsectionhead\par}
            \end{beamercolorbox}
            \vfill
          \end{frame}
        }

        % Footline
        \setbeamertemplate{footline}{
          \hbox{%
          \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,left]{author in head/foot}%
            \hspace{2ex}\usebeamerfont{author in head/foot}\insertshortauthor
          \end{beamercolorbox}%
          \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
            \usebeamerfont{date in head/foot}\insertframenumber{} / \inserttotalframenumber\hspace{2ex}
          \end{beamercolorbox}}%
        }

        % Lists
        \setbeamertemplate{itemize item}{\textcolor{mLightBlue}{$\bullet$}}
        \setbeamertemplate{itemize subitem}{\textcolor{mGray}{$\circ$}}
        \setbeamertemplate{itemize subsubitem}{\textcolor{mGray}{$-$}}
        \setbeamertemplate{enumerate item}{\textcolor{mLightBlue}{\arabic{enumi}.}}
        \setbeamertemplate{enumerate subitem}{\textcolor{mGray}{\alph{enumii}.}}
        \setlength{\itemsep}{0.5em}
        \setlength{\parskip}{0.3em}

        % Blocks
        \setbeamercolor{block title}{fg=white,bg=mLightBlue}
        \setbeamercolor{block body}{fg=mDarkBlue,bg=mLightGray}
        \setbeamercolor{block title alerted}{fg=white,bg=mAccent}
        \setbeamercolor{block body alerted}{fg=mDarkBlue,bg=mAccent!10}

        % Custom text macros (used in slides)
        \newcommand{\keyword}[1]{\textcolor{mLightBlue}{\textbf{#1}}}
        \newcommand{\accenttext}[1]{\textcolor{mAccent}{\textbf{#1}}}
        \newcommand{\highlight}[1]{\textcolor{mGreen}{\textbf{#1}}}
        \newcommand{\important}[1]{\textcolor{mPurple}{\textbf{#1}}}

        % ---- Code blocks: normalsize + extra vertical space via hooks ----
        \usepackage{etoolbox}

        % Default size
        \newcommand{\CodeSize}{\normalsize}
        \newcommand{\SetCodeSize}[1]{\gdef\CodeSize{#1}}
        
        % Hooks for Pandoc's Highlighting environment
        \AtBeginEnvironment{Highlighting}{\vspace{2pt}\CodeSize}
        \AfterEndEnvironment{Highlighting}{\vspace{2pt}\SetCodeSize{\normalsize}}
        
        % Also adjust font used inside
        \makeatletter
        \def\verbatim@font{\CodeSize\ttfamily}
        \makeatother

classoption:
  - compress
  - fragile
execute:
  echo: true
  warning: false
  message: false
  comment: ""
  R.options:
    width: 50
---

```{r setup, include=FALSE}
library(ggplot2)
library(readxl)
library(haven)
library(stargazer)
library(knitr)
library(tidyverse)

# Create a modern theme for plots
theme_modern <- function() {
  theme_minimal(base_size = 10) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = "#E0E0E0", size = 0.3),
      panel.grid.minor = element_blank(),
      text = element_text(family = "sans", color = "#23373b"),
      axis.text = element_text(color = "#4A4A4A"),
      axis.title = element_text(face = "bold", size = 11),
      plot.title = element_text(face = "bold", size = 14, hjust = 0),
      plot.subtitle = element_text(size = 12, hjust = 0, color = "#4A4A4A"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold")
    )
}

theme_set(theme_modern())
set.seed(123)

# Modern color palette
modern_colors <- c("#0066CC", "#EB811B", "#14B03D", "#CC0033", "#6B3AA0", "#F7B500")
scale_color_modern <- function(...) scale_color_manual(values = modern_colors, ...)
scale_fill_modern <- function(...) scale_fill_manual(values = modern_colors, ...)

options(
  width = 50,
  scipen = 999,
  digits = 3
)

opts_chunk$set(
  tidy = FALSE,
  size = "normalsize"
)
```



# Exercises: Background and Tools

### Exercise a: setting up a working environment

```{=tex}
\begin{enumerate}
\tightlist
\item Open \keyword{RStudio} and navigate to your desired working folder

\vspace{0.3em}

\item Create a new folder called \accenttext{r\_course}

\vspace{0.3em}

\item Set it as your \keyword{working directory}

\vspace{0.3em}

\item Create subfolders: \highlight{data} and \highlight{code}

\vspace{0.3em}

\item[(*)] Create a new \important{R Project} in the \texttt{r\_course} folder

\vspace{0.3em}

\item[(*)] Choose your favorite R-studio \important{appearance}: `Tools > Global Options > Appearance`  
\end{enumerate}
```
### Exercise b: r scripts

1.  In the R console, type:

```{r, eval=FALSE}
print("Hello world")
```

2.  Create a new \keyword{R script}: `File > New File > R Script` (or `Ctrl+Shift+N`)

3.  Type the same code in the script and \accenttext{run it} using `Ctrl+Enter` or `Ctrl+As+Enter`


## Exercises II: First Steps and Basic Concepts 

### Exercise a: write a sum function

Write a function that takes a \keyword{numeric vector} as input and returns the \accenttext{sum} of the vector's elements.

```{r, eval=FALSE}
my_sum <- function(x) {
    # Your code here
}
```

Test by comparing with built-in `sum()` function.

### Exercise b: robustness and warnings

Test your `my_sum()` function with:

```{r, eval=FALSE}
numbers2 <- c("1", "2", "3")
```

Add \keyword{error checking} to make the function robust.

### Exercise c: standard deviation function

Implement a function to compute the \accenttext{standard deviation}:

$$\text{SD} = \sqrt{\frac{1}{N-1}\sum_{i=1}^N (x_i - \bar{x})^2}$$

### Exercise d: standard error function

Building on Exercise C, implement:

$$\text{SE}_{\bar{x}} = \frac{\text{SD}}{\sqrt{N}}$$

### Exercise e: t-test

Implement a \keyword{one-sample t-test} function:

$$t = \frac{\bar{x} - \mu_0}{\text{SE}_{\bar{x}}}$$

### Exercise f: fibonacci sequence

Generate the first 30 \accenttext{Fibonacci numbers} where:

-   $F_0 = 0$, $F_1 = 1$
-   $F_n = F_{n-1} + F_{n-2}$ for $n > 1$

### Exercise g: multiples of 3 and 5 (\*)

Write a function that computes the sum of all \keyword{multiples of 3 or 5} up to a number $N$.

\accenttext{Hint:} Multiples of both 3 and 5 should only be added once!

### Exercise h: prime numbers (\*)

Write a function that computes the sum of all \highlight{prime numbers} up to a number $N$.


\normalsize

## Exercise i: Working with Data using dplyr

### Exercise i.1

-   Load the dataset with `data(swiss)`\
-   Use `mutate()` to create two new variables:
    -   `UrbanizationRate`: the inverse of Agriculture (100 – Agriculture)\
    -   `FertilityCategory`: a factor variable with levels:
        -   0–60: "Low"\
        -   60–80: "Medium"\
        -   80–100: "High"\
-   Hint: use `case_when()` for categorization

### Exercise i.2

-   Filter observations where Fertility \> 70\
-   Arrange results by `Education` in descending order

------------------------------------------------------------------------

### Exercise i.3

-   Group data by `FertilityCategory`\
-   Calculate average `Education` for each category using `group_by()` and `summarize()`

### Exercise i.4

-   Find the top 3 observations with highest Fertility values\
-   Display only `Fertility` and `Education` columns

### Exercise i.5

-   Create `EducationToFertilityRatio = Education / Fertility`\
-   Show all observations where the ratio is between 0.1 and 0.2

### Exercise i.6

-   Group by `FertilityCategory`\
-   Calculate:
    -   Median Fertility\
    -   Standard deviation of Education\
-   Store the results in a new data frame

### Exercise i.7

-   Use `n_distinct()` to determine the number of distinct values in:
    -   `Education`\
    -   `FertilityCategory`

## Exercise j: Visualizing Data

### Exercise j.1

-   Load the dataset `swiss.csv` or call `data(swiss)`

### Exercise j.2

-   Print the last 3 observations of the dataset\
-   Create summary statistics for all variables

### Exercise j.3

-   Create new variables:
    -   `UrbanizationRate` = 100 – Agriculture\
    -   `FertilityCategory`:
        -   0–60: "Low"\
        -   60–80: "Medium"\
        -   80–100: "High"\
-   Hint: use `cut()`

### Exercise j.4

-   Create a histogram of `Fertility`
    -   Fill bars in dark blue\
    -   Add a minimal theme

### Exercise j.5

-   Create a boxplot of the variable `Education`

### Exercise j.6

-   Create a scatter plot of `Education` vs. `Fertility`\
-   Add a regression line (linear model)

### Exercise j.7

-   Create scatter plots of `Agriculture` vs. `Examination`\
-   Show them separately for each `FertilityCategory` using `facet_wrap()`

