---
title: "Exercises and Solutions"
subtitle: "A Brief Introduction to Programming with R"
author: "Ana Armendariz & Federica Mascolo"
date: "September 4, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
    theme: cosmo
    code-copy: true
    df-print: kable
execute:
  echo: true
  warning: false
  message: false
  comment: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
```

```{r, eval=FALSE}
install.packages(c("dplyr", "ggplot2"))
library(dplyr)
library(ggplot2)
```

# Part I: Background and tools

### Exercise a: setting up a working environment

1. Open RStudio and navigate to your desired working folder
2. Create a new folder called `r_course`
3. Set it as your working directory
4. Create subfolders: `data` and `code`
5. (\*) Create a new R Project in the `r_course` folder

**Solution**

```{r, eval=FALSE}
dir.create("r_course")
setwd("r_course")
dir.create("data")
dir.create("code")
getwd()
```

R Project: `File > New Project > Existing Directory`, then pick `r_course`.

### Exercise b: R scripts

1. In the R console, type `print("Hello world")`
2. Create a new R script: `File > New File > R Script` (or `Ctrl+Shift+N`)
3. Type the same code in the script and run it using `Ctrl+Enter`

**Solution**

```{r}
print("Hello world")
```

# Part II: First steps and basic concepts

### Exercise a: write a sum function

Write a function that takes a numeric vector as input and returns the sum of the
vector's elements. Test by comparing with built-in `sum()` function.

**Solution**

```{r}
my_sum <- function(x) {
  total_sum <- 0

  for (value in x) {
    total_sum <- total_sum + value
  }

  return(total_sum)
}

numeric_vector <- c(1, 5, 4, 3)
my_sum(numeric_vector)
sum(numeric_vector)
```

### Exercise b: robustness and warnings

Test your `my_sum()` function with `numbers2 <- c("1", "2", "3")`.
Add error checking to make the function robust.

**Solution**

```{r}
my_sum <- function(x) {
  if (!is.numeric(x)) {
    warning("The input is not numeric; attempting to convert it.")
    x <- suppressWarnings(as.numeric(x))
  }

  if (anyNA(x)) {
    stop("The input contains values that cannot be converted to numbers.")
  }

  total_sum <- 0

  for (value in x) {
    total_sum <- total_sum + value
  }

  return(total_sum)
}

numbers2 <- c("1", "2", "3")
my_sum(numbers2)
```

### Exercise c: standard deviation function

Implement a function to compute the standard deviation:

$$
\mathrm{SD} = \sqrt{\frac{1}{N-1}\sum_{i=1}^{N}(x_i-\bar{x})^2}
$$

**Solution**

```{r}
my_sd <- function(x) {
  if (!is.numeric(x)) {
    stop("x must be numeric.")
  }

  n <- length(x)
  if (n < 2) {
    stop("x must contain at least two values.")
  }

  x_bar <- mean(x)
  squared_deviations <- (x - x_bar)^2
  sqrt(sum(squared_deviations) / (n - 1))
}

my_sd(numeric_vector)
sd(numeric_vector)
```

### Exercise d: standard error function

Building on Exercise c, implement:

$$
\mathrm{SE}_{\bar{x}} = \frac{\mathrm{SD}}{\sqrt{N}}
$$

**Solution**

```{r}
my_se <- function(x) {
  my_sd(x) / sqrt(length(x))
}

my_se(numeric_vector)
sd(numeric_vector) / sqrt(length(numeric_vector))
```

### Exercise e: t-test

Implement a one-sample t-test function:

$$
t = \frac{\bar{x}-\mu_0}{\mathrm{SE}_{\bar{x}}}
$$

**Solution**

```{r}
my_t_test <- function(x, mu_0) {
  (mean(x) - mu_0) / my_se(x)
}

my_t_test(numeric_vector, mu_0 = 3)
unname(t.test(numeric_vector, mu = 3)$statistic)
```

### Exercise f: Fibonacci sequence

Generate the first 30 Fibonacci numbers where $F_0 = 0$, $F_1 = 1$, and
$F_n = F_{n-1} + F_{n-2}$ for $n > 1$.

**Solution**

```{r}
fibonacci_sequence <- numeric(30)
fibonacci_sequence[1] <- 0
fibonacci_sequence[2] <- 1

for (i in 3:30) {
  fibonacci_sequence[i] <-
    fibonacci_sequence[i - 1] + fibonacci_sequence[i - 2]
}

fibonacci_sequence
```

### Exercise g: multiples of 3 and 5 (\*)

Write a function that computes the sum of all multiples of 3 or 5 up to a
number $N$. Multiples of both 3 and 5 should only be added once.

**Solution**

```{r}
sum_multiples_3_or_5 <- function(N) {
  if (length(N) != 1 || !is.numeric(N) || is.na(N) || N < 1) {
    stop("N must be one positive number.")
  }

  total <- 0

  for (i in seq_len(floor(N))) {
    if (i %% 3 == 0 || i %% 5 == 0) {
      total <- total + i
    }
  }

  return(total)
}

sum_multiples_3_or_5(100)
```

### Exercise h: prime numbers (\*)

Write a function that computes the sum of all prime numbers up to a number $N$.

**Solution**

```{r}
is_prime <- function(x) {
  if (x < 2) {
    return(FALSE)
  }

  largest_possible_divisor <- floor(sqrt(x))

  if (largest_possible_divisor < 2) {
    return(TRUE)
  }

  !any(x %% seq.int(2, largest_possible_divisor) == 0)
}

sum_primes_up_to <- function(N) {
  if (length(N) != 1 || !is.numeric(N) || is.na(N) || N < 1) {
    stop("N must be one positive number.")
  }

  candidates <- seq_len(floor(N))
  prime_numbers <- candidates[
    vapply(candidates, is_prime, logical(1))
  ]

  sum(prime_numbers)
}

sum_primes_up_to(100)
```

# Part III: Working with data

## Exercise i: working with data using dplyr

### Exercise i.1

- Load the dataset with `data(swiss)`
- Use `mutate()` to create two new variables:
  - `UrbanizationRate`: the inverse of Agriculture (100 - Agriculture)
  - `FertilityCategory`: a factor variable with levels 0-60 "Low", 60-80 "Medium", 80-100 "High"
- Hint: use `case_when()` for categorization

**Solution**

```{r}
data(swiss)

swiss_data <- swiss |>
  mutate(
    UrbanizationRate = 100 - Agriculture,
    FertilityCategory = case_when(
      Fertility <= 60 ~ "Low",
      Fertility > 60 & Fertility <= 80 ~ "Medium",
      Fertility > 80 & Fertility <= 100 ~ "High",
      TRUE ~ NA_character_
    ),
    FertilityCategory = factor(
      FertilityCategory,
      levels = c("Low", "Medium", "High")
    )
  )

head(swiss_data)
```

### Exercise i.2

- Filter observations where Fertility > 70
- Arrange results by `Education` in descending order

**Solution**

```{r}
swiss_data |>
  filter(Fertility > 70) |>
  arrange(desc(Education))
```

### Exercise i.3

- Group data by `FertilityCategory`
- Calculate average `Education` for each category using `group_by()` and `summarize()`

**Solution**

```{r}
swiss_data |>
  group_by(FertilityCategory) |>
  summarize(
    AverageEducation = mean(Education, na.rm = TRUE),
    .groups = "drop"
  )
```

### Exercise i.4

- Find the top 3 observations with highest Fertility values
- Display only `Fertility` and `Education` columns

**Solution**

```{r}
swiss_data |>
  slice_max(Fertility, n = 3, with_ties = FALSE) |>
  select(Fertility, Education)
```

### Exercise i.5

- Create `EducationToFertilityRatio = Education / Fertility`
- Show all observations where the ratio is between 0.1 and 0.2

**Solution**

```{r}
swiss_data |>
  mutate(EducationToFertilityRatio = Education / Fertility) |>
  filter(between(EducationToFertilityRatio, 0.1, 0.2))
```

### Exercise i.6

- Group by `FertilityCategory`
- Calculate median Fertility and standard deviation of Education
- Store the results in a new data frame

**Solution**

```{r}
swiss_stats <- swiss_data |>
  group_by(FertilityCategory) |>
  summarize(
    MedianFertility = median(Fertility, na.rm = TRUE),
    StandardDeviationEducation = sd(Education, na.rm = TRUE),
    .groups = "drop"
  )

swiss_stats
```

### Exercise i.7

- Use `n_distinct()` to determine the number of distinct values in `Education` and `FertilityCategory`

**Solution**

```{r}
swiss_data |>
  summarize(
    DistinctEducation = n_distinct(Education),
    DistinctFertilityCategory = n_distinct(FertilityCategory)
  )
```

## Exercise j: visualizing data

### Exercise j.1

- Load the dataset `swiss.csv` or call `data(swiss)`

**Solution**

```{r}
data(swiss)
swiss_data <- swiss
```

### Exercise j.2

- Print the last 3 observations of the dataset
- Create summary statistics for all variables

**Solution**

```{r}
tail(swiss_data, 3)
```

```{r}
summary(swiss_data)
```

### Exercise j.3

- Create new variables:
  - `UrbanizationRate` = 100 - Agriculture
  - `FertilityCategory`: 0-60 "Low", 60-80 "Medium", 80-100 "High"
- Hint: use `cut()`

**Solution**

```{r}
swiss_data$UrbanizationRate <- 100 - swiss_data$Agriculture

swiss_data$FertilityCategory <- cut(
  swiss_data$Fertility,
  breaks = c(0, 60, 80, 100),
  labels = c("Low", "Medium", "High"),
  include.lowest = TRUE
)

head(swiss_data)
```

### Exercise j.4

- Create a histogram of `Fertility`, fill bars in dark blue, add a minimal theme

**Solution**

```{r}
ggplot(swiss_data, aes(x = Fertility)) +
  geom_histogram(binwidth = 5, fill = "darkblue", color = "white") +
  labs(title = "Distribution of Fertility",
       x = "Fertility", y = "Number of observations") +
  theme_minimal()
```

### Exercise j.5

- Create a boxplot of the variable `Education`

**Solution**

```{r}
ggplot(swiss_data, aes(y = Education)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Distribution of Education", y = "Education") +
  theme_minimal()
```

### Exercise j.6

- Create a scatter plot of `Education` vs. `Fertility`
- Add a regression line (linear model)

**Solution**

```{r}
ggplot(swiss_data, aes(x = Education, y = Fertility)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Fertility and Education",
       x = "Education", y = "Fertility") +
  theme_minimal()
```

### Exercise j.7

- Create scatter plots of `Agriculture` vs. `Examination`
- Show them separately for each `FertilityCategory` using `facet_wrap()`

**Solution**

```{r}
ggplot(swiss_data, aes(x = Agriculture, y = Examination)) +
  geom_point() +
  facet_wrap(~ FertilityCategory) +
  labs(title = "Agriculture and Examination by Fertility Category",
       x = "Agriculture", y = "Examination") +
  theme_minimal()
```
